---
title: "Global multidimensional poverty index"
author: "Elvira Khwatenge"
date: "`r Sys.Date()`"
output: html_document
---


## Introduction

An analysis of the Global Multidimensional Poverty Index (MPI) data for various countries and regions around the world.

## Data Loading and Preparation
```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
library(sf)
library(cluster)
```

Load the dataset
```{r}
data <- read.csv('hdx_hapi_poverty_rate_global.csv', stringsAsFactors = FALSE)
```
Clean the dataset
```{r}
data_cleaned <- data[-1, ]
colnames(data_cleaned) <- c('location_code', 'has_hrp', 'in_gho', 'provider_admin1_name', 'admin1_code', 'admin1_name', 'mpi', 'headcount_ratio', 'intensity_of_deprivation', 'vulnerable_to_poverty', 'in_severe_poverty', 'reference_period_start', 'reference_period_end')
```

Convert numeric columns to appropriate data types
```{r}
numeric_columns <- c('mpi', 'headcount_ratio', 'intensity_of_deprivation', 'vulnerable_to_poverty', 'in_severe_poverty')
data_cleaned[numeric_columns] <- lapply(data_cleaned[numeric_columns], as.numeric)
```

Convert date columns to Date format
```{r}
data_cleaned$reference_period_start <- as.Date(data_cleaned$reference_period_start)
data_cleaned$reference_period_end <- as.Date(data_cleaned$reference_period_end)
```

## Analysis

### Summary Statistics

```{r}
summary_statistics <- summary(data_cleaned[numeric_columns])
kable(summary_statistics)
```

### Top 10 Countries with Highest MPI

```{r}
country_summary <- data_cleaned %>%
group_by(location_code) %>%
summarise(across(all_of(numeric_columns), mean, na.rm = TRUE))
top_10_mpi <- country_summary %>%
arrange(desc(mpi)) %>%
head(10)
kable(top_10_mpi)
ggplot(top_10_mpi, aes(x = reorder(location_code, mpi), y = mpi)) +
geom_bar(stat = "identity", fill = "steelblue") +
coord_flip() +
labs(title = "Top 10 Countries with Highest MPI",
x = "Country Code",
y = "Multidimensional Poverty Index (MPI)") +
theme_minimal()
```

### Correlation Analysis
```{r}
correlation_matrix <- cor(data_cleaned[numeric_columns], use = "complete.obs")
kable(correlation_matrix)
ggplot(data = reshape2::melt(correlation_matrix), aes(x = Var1, y = Var2, fill = value)) +
geom_tile() +
geom_text(aes(label = round(value, 2)), color = "white", size = 3) +
scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
labs(title = "Correlation Heatmap of Poverty Indicators")
```

### Time Series Analysis

```{r}
time_series_data <- data_cleaned %>%
group_by(reference_period_start) %>%
summarise(across(all_of(numeric_columns), mean, na.rm = TRUE))
ggplot(time_series_data, aes(x = reference_period_start, y = mpi)) +
geom_line() +
geom_point() +
labs(title = "Global MPI Trend Over Time",
x = "Year",
y = "Average Multidimensional Poverty Index (MPI)") +
theme_minimal()
```

### Regional Analysis

```{r}
regional_summary <- data_cleaned %>%
group_by(admin1_name) %>%
summarise(across(all_of(numeric_columns), mean, na.rm = TRUE)) %>%
arrange(desc(mpi))
top_20_regions <- head(regional_summary, 20)
kable(top_20_regions)
ggplot(top_20_regions, aes(x = reorder(admin1_name, mpi), y = mpi)) +
geom_bar(stat = "identity", fill = "darkgreen") +
coord_flip() +
labs(title = "Top 20 Regions with Highest MPI",
x = "Region",
y = "Multidimensional Poverty Index (MPI)") +
theme_minimal() +
theme(axis.text.y = element_text(size = 8))
```

## Additional Analyses

### Regional Disparities Analysis

```{r}
regional_summary <- data_cleaned %>%
group_by(admin1_name) %>%
summarise(
mean_mpi = mean(mpi, na.rm = TRUE),
sd_mpi = sd(mpi, na.rm = TRUE),
min_mpi = min(mpi, na.rm = TRUE),
max_mpi = max(mpi, na.rm = TRUE)
)
anova_result <- aov(mpi ~ admin1_name, data = data_cleaned)
summary(anova_result)
```

### Temporal Trend Analysis

```{r}
temporal_trends <- data_cleaned %>%
group_by(location_code, admin1_name) %>%
arrange(reference_period_start) %>%
summarise(
start_mpi = first(mpi),
end_mpi = last(mpi),
years = as.numeric(difftime(last(reference_period_end), first(reference_period_start), units = "days")) / 365,
avg_annual_change = (end_mpi - start_mpi) / years
)
ggplot(data_cleaned, aes(x = reference_period_start, y = mpi, color = admin1_name)) +
geom_line() +
theme_minimal() +
labs(title = "MPI Trends Over Time", x = "Year", y = "MPI", color = "Region")
```

### Decomposition Analysis
```{r}
indicators <- c("mpi", "headcount_ratio", "intensity_of_deprivation")
data_cleaned %>%
gather(key = "indicator", value = "value", indicators) %>%
group_by(admin1_name, indicator) %>%
summarise(mean_value = mean(value, na.rm = TRUE)) %>%
ggplot(aes(x = admin1_name, y = mean_value, fill = indicator)) +
geom_bar(stat = "identity", position = "stack") +
theme_minimal() +
labs(title = "MPI Composition by Region", x = "Region", y = "Contribution to MPI")
```

### Cluster Analysis

```{R}
set.seed(123)
kmeans_result <- kmeans(data_cleaned[c("mpi", "headcount_ratio", "intensity_of_deprivation")], centers = 5)
data_cleaned$cluster <- kmeans_result$cluster
ggplot(data_cleaned, aes(x = headcount_ratio, y = intensity_of_deprivation, color = factor(cluster))) +
geom_point() +
theme_minimal() +
labs(title = "Cluster Analysis of Poverty Profiles", color = "Cluster")
```

### Inequality Analysis
```{r}
gini_coefficients <- data_cleaned %>%
group_by(admin1_name) %>%
summarise(gini_mpi = ineq::Gini(mpi))
ggplot(data_cleaned, aes(x = cumsum(mpi) / sum(mpi), y = seq_along(mpi) / length(mpi))) +
geom_line() +
geom_abline(linetype = "dashed") +
theme_minimal() +
labs(title = "Lorenz Curve of MPI", x = "Cumulative Share of MPI", y = "Cumulative Share of Population")
```

## Conclusion

This is an analysis of the Global Multidimensional Poverty Index data revealing variations in poverty levels across countries and regions, as well as over time..