---
title: "Global Multidimensional Poverty Index Data"
author: "Elvira Khwatenge"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

## Introduction

This report analyzes the Global Multidimensional Poverty Index (MPI) data for various countries and regions around the world.

## Data Loading and Preparation

```{r}
library(dplyr)
library(ggplot2)
library(tidyr)
library(knitr)
library(scales)
```

Load the dataset
```{r}
data <- read.csv("hdx_hapi_poverty_rate_global.csv", stringsAsFactors = FALSE)
```

Clean the dataset
```{r}
data_cleaned <- data[-1, ]
colnames(data_cleaned) <- c('location_code', 'has_hrp', 'in_gho', 'provider_admin1_name', 'admin1_code', 'admin1_name', 'mpi', 'headcount_ratio', 'intensity_of_deprivation', 'vulnerable_to_poverty', 'in_severe_poverty', 'reference_period_start', 'reference_period_end')

```
Convert numeric columns to appropriate data types

```{r}
numeric_columns <- c('mpi', 'headcount_ratio', 'intensity_of_deprivation', 'vulnerable_to_poverty', 'in_severe_poverty')
data_cleaned[numeric_columns] <- lapply(data_cleaned[numeric_columns], as.numeric)

```

Convert date columns to Date format
```{r}
data_cleaned$reference_period_start <- as.Date(data_cleaned$reference_period_start)
data_cleaned$reference_period_end <- as.Date(data_cleaned$reference_period_end)
```

## Analysis

### Summary Statistics

```{r}
summary_statistics <- summary(data_cleaned[numeric_columns])
kable(summary_statistics)
```

### Top 10 Countries with Highest MPI

```{r}
country_summary <- data_cleaned %>%
group_by(location_code) %>%
summarise(across(all_of(numeric_columns), mean, na.rm = TRUE))
top_10_mpi <- country_summary %>%
arrange(desc(mpi)) %>%
head(10)
kable(top_10_mpi)
ggplot(top_10_mpi, aes(x = reorder(location_code, mpi), y = mpi)) +
geom_bar(stat = "identity", fill = "steelblue") +
coord_flip() +
labs(title = "Top 10 Countries with Highest MPI",
x = "Country Code",
y = "Multidimensional Poverty Index (MPI)") +
theme_minimal()

```

### Correlation Analysis

```{r}
correlation_matrix <- cor(data_cleaned[numeric_columns], use = "complete.obs")
kable(correlation_matrix, digits = 3)
ggplot(data = reshape2::melt(correlation_matrix), aes(x = Var1, y = Var2, fill = value)) +
geom_tile() +
geom_text(aes(label = round(value, 2)), color = "white", size = 3) +
scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
theme_minimal() +
theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
labs(title = "Correlation Heatmap of Poverty Indicators")
```

### Time Series Analysis
```{r}
time_series_data <- data_cleaned %>%
group_by(reference_period_start) %>%
summarise(across(all_of(numeric_columns), mean, na.rm = TRUE))
ggplot(time_series_data, aes(x = reference_period_start, y = mpi)) +
geom_line() +
geom_point() +
labs(title = "Global MPI Trend Over Time",
x = "Year",
y = "Average Multidimensional Poverty Index (MPI)") +
theme_minimal()
```
## Conclusion

 an overview of the Global Multidimensional Poverty Index data. 